---
title: "scASfind tutorial"
author: "Yuyao Song"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{scGOclust_vignette}
  %\usepackage[UTF-8]{inputenc}
---


# Identification of cell type-specific alternative splicing events using scASfind

## Example analysis workflow

**Author:**  
Yuyao Song  <ys585@cam.ac.uk>

**Date:**  
May 2023 

## 1. Set up

### 1.1 Install package

```{r}
# to install the latest release from GitHub
# install.packages("devtools")
# library(devtools)
# devtools::install_github("hemberg-lab/scASfind")
library(scASfind)
```

### 1.2 Load packages

```{r}
# load the package scASfind
# also load packages for data manipulation and visualization

library(tibble)
library(dplyr)
library(tidyr)
library(ggplot2)
library(repr)
```

```{r}
# a useful function for setting figure size

library(repr)
fig <- function(width, heigth){
     options(repr.plot.width = width, repr.plot.height = heigth)
}

# alternative load and save function if using .rda format for index storage
# this is to better comply with R package data standards

load_rda_index <- function(index_obj='index_obj.rda'){
load(index_obj)
index_obj@index <-  new(EliasFanoDB)
index_obj@index$loadByteStream(index_obj@serialized)
return(index_obj)
}

save_rda_index <- function(index_obj){
index_obj@serialized <- index_obj@index$getByteStream()
save(index_obj, file = 'index_obj.rda')
# Clear the serialized stream
index_obj@serialized <- raw()
gc()
}
```

### 1.3 Load or create index {.tabset}
#### Load example index

```{r, results='hide',  message=FALSE}
# These are the pre-built indices using the Tasic et al.(2016) dataset
# The dataset have detailed subtype classifications of glutamatergic neurons and GABA-ergic neurons in the mouse primary visual cortex

## recommended load function from scASfind is:
#index <- loadObject("../data/tasic2016_scASfind_index.rds")

## for this vignette
index <- load_rda_index('../data/index_obj.rda')

```

```{r}
# the index is a SCFind class object
class(index)

# the object contains four attributes
slotNames(index)

# the metadata attribute further stores splicing-related information for comprehending the results
## stats stores the mean and SD of the PSI for each node in the index
## node_list stores node related annotations
## diff_cut is a binary matrix to store if the difference of the PSI in a particular pesudobulk and the mean PSI across all pseudobulks is above the cut-off during index building (default 0.2)
## this is to distinguish between PSI = NA (the junction is unquantified), or the PSI of a node is not sufficiently different from dataset mean (PSI ~ mean_PSI) during raw PSI retrieval from index. Both circumstance is treated as 0 in the splicing matrix and compressed during index building

names(index@metadata)
```

```{r}
index@index
```

```{r}
## the above dataset stores splicing events that has a PSI above tissue average
## the below dataset stores splicing events that has a PSI below tissue average

index@datasets
```
#### Create custom index

```{r}
## to build custom scASfind index from scratch, run the create_index.R script in scASfind/scripts
## this will create the below outputs that we load here
```

```{r}
## first read in all the matrix generated from running build_matrix_original.R
load("../data/tasic_test_matrix_above.rda")
load("../data/tasic_test_stats.rda")
load("../data/tasic_test_gene_node_all.rda")
load("../data/tasic_test_diff_cut.rda")
load("../data/tasic_metadata.rda")
```

```{r}
# make sure all pseudobulks are in metadata
nrow(meta[which(!(meta$pseudo_bulk_ID_new %in% colnames(matrix.above))), ]) == 0
```

```{r}
levels(factor(meta$cell_type))
```

```{r}
meta = meta[match(colnames(matrix.above), meta$pseudo_bulk_ID_new), ]
```

```{r}
## make sure that the order of cell in metadata is the same with order of cell in PSI matrix
identical(meta$pseudo_bulk_ID_new, colnames(matrix.above))
```

```{r}
## build the index for spliced-in events, denoted by 'above' i.e. PSI is above dataset average
idx_above <- buildAltSpliceIndex(psival = matrix.above, metadata = meta, dataset.name = 'above', column.label = 'cell_type', qb = 2)
```

```{r}
idx_above <- addIndexMeta(object = idx_above, stats = stats, node_list = node_list, diff_cut = diff_cut)
```

### {-}
## 2. Exploratory analysis

```{r}
# first, check the encoded cell types
cell_types <- cellTypeNames(index)
cell_types
```

```{r}
# check the encoded genes
head(scfindNodes(object = index))
```

### 2.1 Search marker nodes of an interested cell type

```{r}
interested_cell_type <- "above.Astro_Gja1"

# rank the splicing nodes by their frequency of inclusion that is above dataset mean in the interested cell type

markers <- cellTypeMarkers(index, interested_cell_type, background.cell.types = cellTypeNames(index, datasets = 'above'), top.k = 20)
markers
```

```{r}
# plot the nodes using their precision. recall and f1 scores to examine

ggplot(markers, aes(x = precision, y = recall, colour = f1, label = Node_name)) +
theme_bw() + 
geom_point(aes(size = f1)) +
geom_text(vjust = 0, 
          nudge_y = 0.01, 
          colour = "black", 
          check_overlap = T) +
xlim(0.4, 1.2)

fig(6, 6)
```

```{r}
## get the annotations of query nodes
## use the markers of interested cell type as example
nodeDetails(index, as.character(markers$nodes))
```

### 2.2 Search cell types enriched in an interested splicing event

```{r}
# find all nodes in a particular gene

Gria2_nodes <- geneNodes(index, "Gria2", "Gene_name")
Gria2_nodes <- Gria2_nodes[order(as.numeric(as.character(Gria2_nodes$Node))), ]
Gria2_nodes
```

```{r}
# query significant cell types by a specific splicing event
## accept logical operators: prepend "-" means node exclusion

Gria2_mxe <- c("ENSMUSG00000033981.14_28", "ENSMUSG00000033981.14_29")
hyperQueryCellTypes(index, c("ENSMUSG00000033981.14_28", "-ENSMUSG00000033981.14_29"), datasets = 'above') %>% arrange(pval)
```

### 2.3 Retrieve actual PSI for discovered splicing patterns

```{r}
## check the raw PSI value of query nodes in particular cell types, return a dataframe with cell type, node raw PSI and node details

raw_psi = getRawPsi(object = index, node.list = Gria2_mxe, cell.types = gsub("above\\.", "", cellTypeNames(index, datasets = 'above')), verbose = FALSE)
head(raw_psi)
```

### 2.4 Heatmap of PSI value of a splicing pattern

```{r}
## generate a heatmap for raw PSI values

plotRawPsiHeatmap(raw_psi = raw_psi, node.list = Gria2_mxe, cell.types = gsub("above\\.", "", cellTypeNames(index, datasets = 'above')))

fig(6, 10)
```
### 2.5 Find nodes with a similar splicing pattern with query nodes

```{r, results='hide', message=FALSE}
## takes a few seconds to run
similar_nodes = findSimilarNodes(object = index, node.list = Gria2_mxe[1], datasets = 'above', top.k = 10)
```
```{r}
head(similar_nodes)
```

### 2.6 find house keeping nodes in a particular group of cell types

```{r}
above_glul5 = cell_types[grep("above\\.L5", cell_types)]
above_glul5
```


```{r}
## use L5 glutamatergic neurons as example

above_glul5_nodes = findHouseKeepingNodes(object = index, cell.types = above_glul5)
above_glul5_nodes
```

```{r}
# show annotation of splicing nodex
nodeDetails(object = index, node.list = above_glul5_nodes)
```

### 2.7 Find signature splicing events for particular cell types
```{r}

findNodeSignatures(object = index, cell.types = above_glul5)
```

### 2.8 find all cells per cell type that express the full set of nodes in some genes
```{r}

findCellTypes(object = index, gene.list = Gria2_mxe[1], datasets = 'above')
```


### 2.9 long query optimization routine

```{r}
## look at all nodes in Gria2 gene, this is what we want to shorten to find nodes that are cell type specific
Gria2_nodes$Node_id
```

```{r}
# long query optimization function, finds node sets that are significantly enriched in some cell types
Gria2_markers = markerNodes(object = idx_above, node.list = Gria2_nodes$Node_id, log.message = FALSE)
head(Gria2_markers)
```

```{r}
# get coordinated nodes in a gene
getCoordinatedNodes(object = idx_above, gene.name = 'Gria2')
```

```{r}
# another function to find cell type specific node sets, return cell types enriched in the patterns found
findNodeSets(object = idx_above, gene.list = c("Golgb1"), query.type = 'Gene_name', node.types = c("CE", "AA", 'AD', 'RI', NA))
```


```{r}
sessionInfo()
```

```{r}
rm()
```

